---
title: "US Economic Productivity and Minimum Wage"
author: "Jane Huber & Meghan Mokate"
date: "4/30/2021"
output:
  html_document: default
---

## Introduction

Minimum Wage is a piece of federal economic policy that sets a floor for wage work in the United States. Originally brought into law during the Great Depression in the 1930’s, the aim was to create “minimum standard of living necessary for health, efficiency and general well-being, without substantially curtailing employment”.

Ever since its creation, there has been vigorous debate across the US political spectrum as to what the value of a minimum wage should be or there should be one at all. As the “Fight for 15” political movement rises to prominence on the national level, some economists argue that $15/hour simply is not ambitious enough or a reflection of the economy as a whole. The Center for Economic and Policy Research (CEPR) claims, “Until 1968, the minimum wage not only kept pace with inflation, it rose in step with productivity growth. The logic is straightforward; we expect that wages in general will rise in step with productivity growth. For workers at the bottom to share in the overall improvement in society’s living standards, the minimum wage should also rise with productivity.”

We will examine that claim through our own analysis of the numbers, looking to see what the value of the minimum wage would be if, from its inception, minimum wage was increased according to those two variables. 

## Findings

--> Need a quick summary of findings.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)

library(tidyverse) 
library(rvest)
library(janitor)
library(tinytex)
library(knitr)
library(splitstackshape) # for cSplit
library(extrafont)
library(lubridate) # to help clean date representations
library(kableExtra)
library(formattable) # to change value to currency
```
```{r gather-clean-productivity-data, include = FALSE}

#Source: https://www.epi.org/productivity-pay-gap/
net_productivity_dataset <- read_csv("Data_HourlyCompensation_NetProd.csv")

net_productivity_dataset <- 
  net_productivity_dataset %>% 
  select(Year, `Net Productivity`) %>% 
  transmute(year = Year,
            net_productivity = `Net Productivity`)

```
```{r gather-clean-minimum-wage, include = FALSE}
#Source: https://www.dol.gov/agencies/whd/minimum-wage/history/chart#fn4
min_wage_dataset <- read_csv("FederalMinimumWage.csv")

min_wage_dataset <- 
  min_wage_dataset %>% 
  transmute(year = as.numeric(format(as.Date(mdy(effective_date)), "%Y")),
            raw_min_wage = 
              as.numeric(str_replace_all(fed_min_wage, "\\$", "")))

# NOTE: At this point, the minimum wage dataset only contains years in which
# the minimum wage was adjusted by Congress. We will address this as we
# combine datasets.

```
```{r gather-clean-cpi-data, include = FALSE}

# Get "seriesId" from BLS website.This refers to the CPI for all items
# for urban consumers, which is considered the standard for economic
# analysis.
# 
# bls_seriesid <- "CUSR0000SA0"
# 
# # BLS API only allows for 10 years of data to be gathered at a time.
# # Therefore, calculate the number of iterations through a for loop
# # needed to capture all data.
# 
# num_decades <- round((2018 - 1948) / 10) + 1
# 
# # Set initial values for start and end year. In the for loop, we will update
# # these values after gathering the data.
# start_year <- 1948
# end_year <- 1957
# 
# 
# # Declare inflation_dataset in Global context, update in for-loop
# inflation_dataset
# 
# for (i in 1:num_decades) {
# 
#   if (i != 1) {
# 
#     start_year <- end_year + 1
#     end_year <- start_year + 9
# 
#     if (as.numeric(end_year) > 2018) {
#       #This is the final year that we want in the dataset.
#       end_year <- 2018
#     }
#   }
# 
#   print(start_year)
#   print(end_year)
# 
#   decade_dataset <- bls_api(seriesid = bls_seriesid,
#                        startyear = as.character(start_year),
#                        endyear = as.character(end_year),
#                        annualaverage=TRUE)
#   if (i != 1) {
#     inflation_dataset <- bind_rows(inflation_dataset, decade_dataset)
#   } else {
#     inflation_dataset <- decade_dataset
#   }
# }

# write_csv(inflation_dataset, "Inflation_Dataset")

inflation_dataset <- read_csv("Inflation_Dataset")

# The BLS provides us with monthly CPI values for every year since 1948. We 
# need to average these values to create an annual CPI value.

inflation_dataset <- 
  inflation_dataset %>% 
  group_by(year) %>% 
  summarize(annual_avg_cpi = mean(value))

# Note: In order to calculate inflation between two years, we will use the
# following formula (source: BLS):
#   inflation = ((SecondPeriodCPI - PreviousPeriodCPI) / PreviousPeriodCPI)

```
```{r gather-presidents-dataset, include = FALSE}
presidents <- read_csv('presidents.csv') %>%
  cSplit("President Name,Years In Office,Number,Party", ",") %>% # split values into separate columns
  rename(president = "President Name,Years In Office,Number,Party_1", # rename columns
         years = "President Name,Years In Office,Number,Party_2",
         number = "President Name,Years In Office,Number,Party_3",
         party = "President Name,Years In Office,Number,Party_4") %>%
  separate(years, c("start_year", "end_year"), sep = "-") %>% # separate years values
  mutate(end_year = ifelse(president == "William H. Harrison", 1841, end_year)) %>% # fill in some misc missing values 
  mutate(end_year = ifelse(president == "Donald Trump", 2021, end_year)) %>%
  pivot_longer(cols = c(start_year, end_year), # pivot so that years are a column
               names_to = "start_end",
               values_to = "year",
               values_transform = list("year"=as.numeric)) %>%
  select(1, 3, 5) # select only rows we want
```
```{r combine-datasets, include = FALSE}
#First, combine inflation and min_wage datasets
main_dataset <- 
  full_join(min_wage_dataset, inflation_dataset)

#Now, add in the net productivity
main_dataset <- 
  full_join(main_dataset, net_productivity_dataset)

```
```{r tidy-main_dataset, include = FALSE}
# ensure that year is arrange chronologically and fill minimum wage values down so that all years have a value for minimum wage, not just the year it changed
# filter year since our scope is only for 1948 through 2018
main_dataset <- 
  main_dataset %>% 
  arrange(year) %>% 
  fill(raw_min_wage, .direction = "down") %>% 
  filter(year >= 1948)

```
```{r calculate-year-to-year-productivity-gains, include = FALSE}
# We will be using a for loop to calculate this column

# need to create the column with the specified type and length
main_dataset$prod_multiplier <- vector("numeric", nrow(main_dataset))

# loops through the main dataset and takes the difference in percentage year to year, divides this by 100 since it is a percentage, and adds 1, since it is a multiplier
for (i in 1:nrow(main_dataset)) {
  
  main_dataset$prod_multiplier[i] <-  {
    if (i == 1) {
      1
    } 
    else {
      1 + ((main_dataset$net_productivity[i] -
                          main_dataset$net_productivity[i-1]) / 100)
    }
  }
}

```
```{r convert-minwage-to-2018-dollars, include = FALSE}
main_dataset$min_wage_2018_adjusted <- vector("numeric", nrow(main_dataset))

# pull the value for annual average cpi in 2018 and store it in this variable
cpi_for_2018 <- main_dataset[[71,3]]

# loops through rows in main dataset and annual average cpi to a new variable "cpi_for_year"
# creates new variables for a base that assigns the raw minimum wage to the first row (1948) and the minimum wage adjusted for the above year for every subsequent row 
# creates new variables for inflation rate that uses the formula for inflation - (CPIx+1 – CPIx ) / CPIx - and multiplies this by the base minimum for the 1948 row, and finds the year over year percent increase in CPI through taking CPI per year over prior year average CPI and finding adjusted minimum wage through multiplying this percent by the base year minimum wage

for (i in 1:nrow(main_dataset)) {
  
  cpi_for_year <- main_dataset$annual_avg_cpi[i]

  # Get previous minimum wage value, which we will then multiply
  # by the inflation percentage between the two years.
  base_min_wage_for_year <- ifelse(i == 1,
                                   main_dataset$raw_min_wage[i],
                                   main_dataset$min_wage_2018_adjusted[i-1])
  
  if (i == 1) {
  # We use the formula to get inflation and turn it into a decimal.
  # Formula: (CPIx+1 – CPIx ) / CPIx, where:
  # CPIx is Consumer Price Index of Initial Year
  
  inflation_rate <- ((cpi_for_2018 - cpi_for_year) / cpi_for_year)
  
  # Multiply the base minimum wage rate (what it was in previous year) by
  # the inflation rate to get an inflation-adjusted value.
  main_dataset$min_wage_2018_adjusted[i] <- inflation_rate * base_min_wage_for_year
  
  } else {
  
    perc_cpi_increase_year_over_year <- cpi_for_year /
      main_dataset$annual_avg_cpi[i - 1]
    
    main_dataset$min_wage_2018_adjusted[i] <- perc_cpi_increase_year_over_year *
      base_min_wage_for_year
  
  }
}

```
```{r calculate-purchasing-power, include = FALSE}
# calculate purchasing power
main_dataset$cpi_change_from_2018 <- (main_dataset$annual_avg_cpi / cpi_for_2018) * 100
# 1 dollar today = 10 cents in 1948
# money in 1948 had 90% less purchasing power

#applying purchasing power
main_dataset$min_wage_purchasing_power <- (cpi_for_2018 / main_dataset$annual_avg_cpi) * main_dataset$raw_min_wage
# minimum wage was the highest as far as purchasing power goes in the 1970's at ~$10. 

```
```{r calculate_productivity-minwage, include = FALSE}

# Now that minimum wage is in 2018 dollars, we multiple it by raw percentages
# to get the productivity-and-inflation-based minimum wage. All is in 2018
# dollars.

# need to create the column with the specified type and length
main_dataset$min_wage_prod_adjusted <- vector("numeric", nrow(main_dataset))

# loop through each row in the dataset to return adjusted minimum wage multiplied by the productivity multiplier for 1948 and productivity adjusted minimum wage multiplied by the productivity multiplier for all other years
for (i in 1:nrow(main_dataset)) {

  main_dataset$min_wage_prod_adjusted[i] <- 
    ifelse(i==1, 
           (main_dataset$min_wage_2018_adjusted[1] * 
             main_dataset$prod_multiplier[i]),
           (main_dataset$min_wage_prod_adjusted[i-1] *
                          main_dataset$prod_multiplier[i]))
  
}

```
```{r update-dataset-only-necessary-columns, include = FALSE}

# Here, we remove unnecessary columns and pivot longer to facilitate graphing
# our data

main_dataset_final <- 
  main_dataset %>%
  pivot_longer(cols = c(raw_min_wage, 
                        min_wage_2018_adjusted,
                        min_wage_prod_adjusted,
                        min_wage_purchasing_power),
               names_to = "Type",
               values_to = "Dollars") %>% 
  select(year, "Type", "Dollars") # only want several columns relevant for the graph

```
```{r create-theme, include = FALSE}
# 1. Includes the minimal theme included in R
# 2. Assigns the Georgia font to graph text
# 3. Specifies that the title is bold, size 16, and seagreen
# 4. Specifies that the subtitle is italic, size 14, and royalblue4
# 5. Specifies that the axis text is size 12 and navyblue
# 6. Eliminate axis lines and ticks
# 7. Rotate x axis values 90 degrees so that entire value is visible and axis is cleaner
# 8. Specifies that the legend title is bold, size 12, and seagreen
# 9. Specifies that the legend text is size 12 and navyblue

wage_theme <-
  theme_minimal() + 
  theme(
    #text = element_text(family = "Georgia"), 
    plot.title = element_text(face = "bold", size = 16, color = "seagreen"), 
    plot.subtitle = element_text(face = "italic", size = 14, color = "royalblue4"), 
    axis.title = element_text(size = 14, face = "bold", color = "seagreen"), 
    axis.text = element_text(size = 12, color = "navyblue"), 
    axis.line = element_blank(), 
    axis.ticks = element_blank(), 
    axis.text.x = element_text(angle = 90), 
    legend.title = element_text(size = 12, face = "bold", color = "seagreen"), 
    legend.text = element_text(size = 12, color = "navyblue")) 
```
```{r add-wage-change data, include = FALSE}
# need to create the column with the specified type and length
main_dataset$wage_change <- vector("numeric", nrow(main_dataset))

# loops through rows in main_dataset and returns 0 if they did not change from the previous year (row) and 45 if they did change
for (i in seq(nrow(main_dataset))) {
  if (i == 1) {
    main_dataset$wage_change[i] <- 0
  } else {
    if (main_dataset$raw_min_wage[i] == main_dataset$raw_min_wage[i - 1]) {
    main_dataset$wage_change[i] <- 0 # do not want these values to show up when graphed
    } 
    else { 
    main_dataset$wage_change[i] <- 45 # assign value of 45 so that it will align with our other graph when mapped
    }
  }
}

```
```{r create-graph-comparing-min-wages, echo = FALSE}
wage_values_graph <- 
  ggplot() +
    geom_bar(data = main_dataset, aes(x = year, y = wage_change), alpha = .3, stat = "identity", color = "darkseagreen1") + # shows years that minimum wage changed in green with an alpha to make the bars transparent
  geom_path(data = main_dataset_final, aes(x = year, y = Dollars, color = Type), size = 1.5) + # shows values from the "Type" column of raw_min_wage, min_wage_2018_adjusted, min_wage_prod_adjusted, and min_wage_purchasing_power
  labs(title = "$ Disparity in Actual and Calculated \nMinimum Wage over Time", # title graph
       y = "Dollars", # label y axis
       x = "Year", # label x axis
       caption = "Year when the federal minimum wage changed shown in green") + # add caption to specify what the bar graph is 
  scale_color_discrete(name="Minimum Wage", # label color legend
    labels=c( # change the legend text values
      "2018 GDP Adjusted", 
      "2018 Productivity Adjusted",
      "Purchasing Power",
      "Raw Value")) +
  wage_theme + # add in our theme (created above)
  theme(plot.title = element_text(hjust = 0.5)) # center the title

wage_values_graph # view the graph
```

### 2018 GDP Adjusted Versus Raw Value:

The graph shows the steady shallow slope of the raw value of minimum wage over time, ending in a value of \$7.25 in 2018. 
On the other hand, the 2018 GDP Adjusted line, which calculates minimum wage in 2018 dollars adjusted for GDP, shows a much steeper slope. The slope particularly appears to begin its sharp increase in the late 1970s and really takes off around 1981 (the first year of the Reagan administration), ending with a value of $39.45. The gap between these lines demonstrates the amount that wage has failed to increase proportionately with GDP, or net productivity. In terms of informing policy, a GDP adjusted minimum wage of almost \$40 makes the "Fight for 15" appear much less radical.

### Purchasing Power Versus Raw Value

The graph also shows what the Purchasing Power of minimum wage was over the years in 2018 values. Since the calculation is based on 2018 values, the Raw Value for minimum wage in 2018 is logically equal to its 2018 purchasing power. Though, when looking back year over year, it is clear that the purchasing power of minimum wage has historically been much higher, with the 1940 value of \$4.18 representing a purchasing power of **over ten times** the minimum wage value. 

These metrics highlight that minimum wage over time has failed to keep pace with the economic factors of productivity and purchasing power. 

### Years with Changes in Federal Minimum Wage

Another informative component of the graph is the green bars, which represent years when federal minimum wage increased. From this analysis, we can see that the longest periods where minimum wage went unchanged were from 1982-1989 (Reagan), 1997-2007 (Clinton and Bush), and from 2009-present (Obama, Trump, Biden). Further, 1974-1981 had almost annual increases in federal minimum wage. Though the increases do not appear to have kept pace with the increasing GDP those years, the increases prevented the gap from widening as much as it would have if the wage had flat-lined. The context provided from viewing historical increases in minimum wage make it clear that the unchanged federal minimum wage from 2009-2018 (and, indeed, to 2021), is an uncommonly long duration. Further, while raw minimum wage has plateaued, 2018 GDP Adjusted values continue to increase sharply and steadily. 

## Snapshot of Minimum Wage Increases

While reflecting on patterns of minimum wage value and increases, it can be helpful to take a snapshot of how minimum wage looks in years in which Congress voted to increase that value.  Alongside the raw minimum wage value, the purchasing power of minimum wage, and a productivity-increased minimum wage value, we wanted to include which president oversaw the increase in minimum wage and the number of years since the previous change.

We wanted to add these values in context because it allows for US observers to see more political context when thinking about minimum wage changes and, in particular, to emphasize the lack of steady and regular increases in the minimum wage in the second half of the 20th Century and the beginning of the 21st Century.

It is exactly this lack of increases in the minimum wage that has contributed to the lessening purchasing power of minimum wage, which contributes to many economic issues in the country such as widening inequality.

```{r adjust-presidents-dataset, include = FALSE}
# Add in president data pulled in earlier
# This helps us to provide some analysis of what administrations increased wages for context, referenced in our interpretation of our chart

wage_with_pres <-
  left_join(main_dataset, presidents) %>% 
  fill(president, .direction = "up") %>%  # ensure that all years have president values
  filter(!duplicated(year, fromLast = T))

```
```{r create-table-dataset, include = FALSE}
# Create table data; we want to narrow down
table_dataset <- 
  left_join(wage_with_pres, min_wage_dataset) %>% 
  filter(wage_change == 45) %>%
  filter(year >= 1948) %>% 
  select(year,
         president,
         raw_min_wage,
         min_wage_purchasing_power,
         min_wage_2018_adjusted,
         min_wage_prod_adjusted) %>%
  mutate("Number of Years Since Last Change" = year - lag(year, default = first(year))) %>% # calculate how long it has been since the last increase in minimum wage
  mutate(year = as.character(year))

# Convert all dollar values to currency type
table_dataset$raw_min_wage <- currency(table_dataset$raw_min_wage, digits = 2L)
table_dataset$min_wage_purchasing_power <- currency(table_dataset$min_wage_purchasing_power, digits = 2L)
table_dataset$min_wage_2018_adjusted <- currency(table_dataset$min_wage_2018_adjusted, digits = 2L)
table_dataset$min_wage_prod_adjusted <- currency(table_dataset$min_wage_prod_adjusted, digits = 2L)
```
```{r organize-table-data, include = FALSE}

long_gap_min_wage_change <- ifelse(table_dataset$`Number of Years Since Last Change` > 5, "red", "black")

```
```{r create-table, echo = FALSE}


year_footnote <- "Year congress updated minimum wage."
raw_min_wage_footnote <- "Original dollar amounts."
purchasing_power_footnote <- "Equivalent to in 2018 dollars."
actual_min_wage_footnote <- "Minimum wage if it was adjusted for inflation."
productivity_min_wage_footnote <- "Minimum wage if increased along with productivity."
num_years_since_change_footnote <- "We highlight years where it has been more than 5 years since a minimum wage change."

footnotes <- c(year_footnote,
               raw_min_wage_footnote,
               purchasing_power_footnote,
               actual_min_wage_footnote,
               productivity_min_wage_footnote,
               num_years_since_change_footnote)

table <- 
  table_dataset %>% 
  kable(col.names = c("Year[note]", 
                    "President",
                    "Raw Minimum Wage[note]", 
                    "Purchasing Power[note]",
                    "Actual Minimum Wage[note]",
                    "Productivity Minimum Wage[note]",
                    "Number of Years Since Last Change[note]"),
      align = "ccccccc",
      caption = "Minimum Wage",
      format.args = list(big.mark = ","),
      digits = c(0, 2, 2, 2, 2)) %>% 
  kable_paper(font_size = 12,       
              html_font = "Georgia",
              full_width = T) %>% 
  kable_styling(latex_options = "striped") %>% 
  add_footnote(footnotes) %>%
  row_spec(0, bold = T, color ="seagreen", font_size = 15) %>% #Add styling to header
  column_spec(7, color = long_gap_min_wage_change) %>% # change font to red in red years that have not had a minimum wage increase in a long time
  column_spec(1, bold = T) 

table

```

## Appendix I: Methodology and Data

As discussed in the introduction, the current discussions surrounding minimum wage increases are very polarizing, with impassioned beliefs of pros and cons from proponents and opponents of the measure. Key arguments made against raising minimum wage are that the increase could lead to inflation and job losses, and negatively impact businesses. Conversely, proponents raise points about the increased costs of living and claim that the current rate is not sufficient income.

Independent of these arguments, our calculations seek to mathematically assess how minimum wage over time has related to important economic factors: purchasing power and worker productivity. In doing this, we can provide color about what how minimum wage has changed contextually from 1940 to 2018. The findings of our calculations are visually presented through the graph, which allows us to easily identify how actual raw minimum wage relates to the identified economic factors year over year.

### Gather Datasets

<li>Data Sources</li>
- Productivity data from: https://www.epi.org/productivity-pay-gap/
  - Note that this data comes from a policy research institute, the Economic Policy Institute, but its source is the Bureau of Labor Statistics (BLS) Labor Productivity and Costs program.
- Minimum wage data from: https://www.dol.gov/agencies/whd/minimum-wage/history/chart#fn4
- CPI data from the Bureau of Labor Statistics that details the annual average CPI for the years in our scope. Data download is made possible via an API.
    - In order to calculate the purchasing power of a dollar in 1950 in 2018 dollars, the conversion formula is: ((CPI2 - CP1) / CP1) * 100. CPI2 = CPI value in the later year; CPI1 = CPI value of the earlier year. That value would then be multiplied by the dollar amount in 1950 that we seek to convert in 1950.
    - Useful Q & A about what CPI is and how to interpret: https://www.bls.gov/cpi/questions-and-answers.htm#Question_15 </liii>
- Presidents data from: ________

These source of all of these data are government sources and are therefore considered reliable. There are no obvious limitations to the data, except that perhaps the productivity data was cleaned by a third party and a mistake could have been made. That said, we have no reason to be directly concerned about that data.

### Gather and Clean Data

Guide to key datasets:
1. main_set = includes all calculations and multiplies and is un-pivoted (wide)
2. main_dataset_final = includes all calculations and multiplies and is pivoted (long) by dollar value calculations each year
3. wage_values_graph = graphed findings of cacluations
4. table = table for values in years when minimum wage changed

### Calculations

Productivity Multiplier: [add more context, explain method]
The productivity multiplier plays a very important role in how we understand the value of what minimum wage would be if it rose alongside GDP. 


Purchasing Power of Minimum Wage: [add more context, explain method]
Here, we want to convert all dollar amounts in our dataset to 2018 dollars.
This will make all the values more readable to our audience.

Create column that divides the average CPI each year by the CPI for 2018 (found above) and multiples by 100

Value of Minimum Wage Factoring in Productivity: [add more context, explain method]
Calculate the productivity-and-inflation-based minimum wage in 2018 dollars

### Final adjustments to data for graph and table

Since our datasets were relatively clean and tidy, there were not many changes needed to use them for the graph and table that we present.

Graph Adjustments:
That said, our main_dataset needed to be adjusted ...

Table Data Adjustments:
