---
title: "HuberMokate_22April"
author: "Jane Huber"
date: "4/22/2021"
output: html_document
---

The point of this document is to share our work with Alea/Prof Fusi.
Contains our work, not much more.

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) 
library(rvest)
library(janitor)
library(tinytex)
library(haven) # may use for regression / regression line (may not)
library(splitstackshape) # for cSplit
library(knitr)
library(extrafont)
library(blscrapeR) # to gather and complete inflation-related calculations 
library(lubridate) # to help clean date representations
```

## Gather and clean original data

```{r gather-clean-productivity-data}

#Source: https://www.epi.org/productivity-pay-gap/
net_productivity_dataset <- read_csv("Data_HourlyCompensation_NetProd.csv")

net_productivity_dataset <- 
  net_productivity_dataset %>% 
  select(Year, `Net Productivity`) %>% 
  transmute(year = Year,
            net_productivity = `Net Productivity`)

```

```{r gather-clean-minimum-wage}
#Source: https://www.dol.gov/agencies/whd/minimum-wage/history/chart#fn4
min_wage_dataset <- read_csv("FederalMinimumWage.csv")

min_wage_dataset <- 
  min_wage_dataset %>% 
  transmute(year = as.numeric(format(as.Date(mdy(effective_date)), "%Y")),
            raw_min_wage = 
              as.numeric(str_replace_all(fed_min_wage, "\\$", "")))

# min_wage_dataset %>% 
#   complete(year)

# Transmute year, we have to do some work to parse out the year value
# Transmute raw_min_wage, we want to remove the $ sign.

# NOTE: At this point, the minimum wage dataset only contains years in which
# the minimum wage was adjusted by Congress. We will address this as we
# combine datasets.

```

Useful Q & A about what CPI is and how to interpret:
https://www.bls.gov/cpi/questions-and-answers.htm#Question_15
```{r gather-clean-cpi-data}

# Get "seriesId" from BLS website.This refers to the CPI for all items
# for urban consumers, which is considered the standard for economic
# analysis.
# 
# bls_seriesid <- "CUSR0000SA0"
# 
# # BLS API only allows for 10 years of data to be gathered at a time.
# # Therefore, calculate the number of iterations through a for loop
# # needed to capture all data.
# 
# num_decades <- round((2018 - 1948) / 10) + 1
# 
# # Set initial values for start and end year. In the for loop, we will update
# # these values after gathering the data.
# start_year <- 1948
# end_year <- 1957
# 
# 
# # Declare inflation_dataset in Global context, update in for-loop
# inflation_dataset
# 
# for (i in 1:num_decades) {
# 
#   if (i != 1) {
# 
#     start_year <- end_year + 1
#     end_year <- start_year + 9
# 
#     if (as.numeric(end_year) > 2018) {
#       #This is the final year that we want in the dataset.
#       end_year <- 2018
#     }
#   }
# 
#   print(start_year)
#   print(end_year)
# 
#   decade_dataset <- bls_api(seriesid = bls_seriesid,
#                        startyear = as.character(start_year),
#                        endyear = as.character(end_year),
#                        annualaverage=TRUE)
#   if (i != 1) {
#     inflation_dataset <- bind_rows(inflation_dataset, decade_dataset)
#   } else {
#     inflation_dataset <- decade_dataset
#   }
# }

# write_csv(inflation_dataset, "Inflation_Dataset")

inflation_dataset <- read_csv("Inflation_Dataset")

# The BLS provides us with monthly CPI values for every year since 1948. We 
# need to average these values to create an annual CPI value.

inflation_dataset <- 
  inflation_dataset %>% 
  group_by(year) %>% 
  summarize(annual_avg_cpi = mean(value))

# Note: In order to calculate inflation between two years, we will use the
# following formula (source: BLS):
#   inflation = ((SecondPeriodCPI - PreviousPeriodCPI) / PreviousPeriodCPI)

```

## Combine and Clean Data

```{r combine-datasets}

#First, combine inflation and min_wage datasets
main_dataset <- 
  left_join(inflation_dataset, min_wage_dataset, by = "year")
  

#Now, add in the net productivity
main_dataset <- 
  left_join(main_dataset, net_productivity_dataset, by = "year")

```

```{r tidy-main_dataset}

main_dataset <- 
  main_dataset %>% 
  fill(raw_min_wage, .direction = "down") 


# In 1948, the minimum wage was $.40. Since our current method of filling in NA
# values, relies on that value, 
main_dataset$raw_min_wage <- 
  replace_na(main_dataset$raw_min_wage,  0.4) #How can we get this to work 
# in the pipes?

```

Issue:
Not sure if this multiplier is the right approach. The aim is to capture 
productivity increase and factor that in to value of minimum wage.

```{r calculate-year-to-year-productivity-gains}

main_dataset$prod_multiplier <- vector("numeric", nrow(main_dataset))

for (i in 1:nrow(main_dataset)) {
  
  if (i == 1) {
    multiplier <- 1
  } else if (i == 2) {
    multiplier <- main_dataset$net_productivity[i]
  } else {
    multiplier <- main_dataset$net_productivity[i] /
                        main_dataset$net_productivity[i-1]
  }
  
  main_dataset$prod_multiplier[i] <- multiplier
  
}

```

## Create calculations

Issue: We are using the inflation formula, but the data still doesn't seem right.
Things should start at ~$4, I expect it to go up to like $13.

```{r convert-minwage-to-2018-dollars}

# Here, we want to convert all dollar amounts in our dataset to 2018 dollars.
# This will make all the values more readable to our audience.

main_dataset$min_wage_2018_adjusted <- vector("numeric", nrow(main_dataset))

cpi_for_2018 <- main_dataset[[71,2]]

for (i in 1:nrow(main_dataset)) {

  cpi_for_year <- main_dataset$annual_avg_cpi[i]

  # Get previous minimum wage value, which we will then multiply
  # by the inflation percentage between the two years.
  base_min_wage_for_year <- ifelse(i == 1,
                                   main_dataset$raw_min_wage[i],
                                   main_dataset$min_wage_2018_adjusted[i-1])
  
  
  if (i == 1) {
  # We use the formula to get inflation and turn it into a decimal.
  # Formula: (CPIx+1 – CPIx ) / CPIx, where:
  # CPIx is Consumer Price Index of Initial Year
  
  inflation_rate <- ((cpi_for_2018 - cpi_for_year) / cpi_for_year)
  
  # Multiply the base minimum wage rate (what it was in previous year) by
  # the inflation rate to get an inflation-adjusted value.
  main_dataset$min_wage_2018_adjusted[i] <- inflation_rate * base_min_wage_for_year
  } else {
  
    perc_cpi_increase_year_over_year <- cpi_for_year /
      main_dataset$annual_avg_cpi[i - 1]
    
    main_dataset$min_wage_2018_adjusted[i] <- perc_cpi_increase_year_over_year *
      base_min_wage_for_year
    
  }
}

```

```{r calculate_productivity-minwage}

# Now that minimum wage is in 2018 dollars, we multiple it by raw percentages
# to get the productivity-and-inflation-based minimum wage. All is in 2018
# dollars.

main_dataset$min_wage_prod_adjusted <- vector("numeric", nrow(main_dataset))

for (i in 1:nrow(main_dataset)) {

  base_amount <- ifelse(i==1, 
                        main_dataset$min_wage_2018_adjusted[1],
                        main_dataset$min_wage_prod_adjusted[i-1])
  
  main_dataset$min_wage_prod_adjusted[i] <- 
    base_amount * main_dataset$prod_multiplier[i]
  
}

```


## Graph


```{r update-dataset-only-necessary-columns}

# Maybe here, we can remove unnecessary stuff

main_dataset_final <- 
  main_dataset %>%
  select(year, raw_min_wage, min_wage_2018_adjusted, min_wage_prod_adjusted) %>% 
  pivot_longer(cols = c(raw_min_wage, 
                        min_wage_2018_adjusted,
                        min_wage_prod_adjusted),
               names_to = "Type",
               values_to = "Dollars")

```


Create a theme
```{r create-theme}
wage_theme <-
  theme_minimal() + 
  theme(
    # text = element_text(family = "Georgia"), 
    plot.title = element_text(face = "bold", size = 16, color = "seagreen"), 
    plot.subtitle = element_text(face = "italic", size = 14, color = "royalblue4"), 
    axis.title = element_text(size = 14, face = "bold", color = "seagreen"), 
    axis.text = element_text(size = 12, color = "navyblue"), 
    axis.line = element_blank(), 
    axis.ticks = element_blank(), 
    axis.text.x = element_text(angle = 90), 
    legend.title = element_text(size = 12, face = "bold", color = "seagreen"), 
    legend.text = element_text(size = 12, color = "navyblue")) 
```


```{r create-graph-comparing-min-wages}

wage_values_graph <- 
  main_dataset_final %>% 
  ggplot() +
  geom_path(aes(x = year, y=Dollars, color= Type)) +
  labs(title = "$ Disparity in Actual and Calculated Minimum Wage Changes") +
  wage_theme

wage_values_graph

```



# PART TWO - IMPACTFUL FACTORS

## Regression Datasets and Analysis

Create dummy variable for when there was a change in minimum wage
```{r}
regress_set <- master_set %>%
  mutate(change = ifelse(increase_act == 0, 0, 1))
```

## Add in regression datasets: 
(1) unemployment rate
(2) president / party
(3) population / population growth

Unemployment dataset; source: https://data.bls.gov/pdq/SurveyOutputServlet
```{r}
unemploy <- read_csv('unemploy.csv') %>%
  row_to_names(row_number = 11) %>% # make values in row 11 the column titles
  mutate_if(is.character,as.numeric) %>% #change all values to numeric
  rowwise() %>% # prepare to take mean of values
  mutate(avg_unemploy = mean(c_across(2:13))) %>% # create column of mean to get avg_unemploy
  select(1, 14) %>% # select only year and average column
  mutate(avg_unemploy = (avg_unemploy / 100)) # make this into a percentage

regress_set_a <- left_join(regress_set, unemploy, by="Year") # add to dataset
regress_set_a
```

Presidential party for each year dataset; source: https://github.com/awhstin/Dataset-List/blob/master/presidents.csv
```{r}
presidents <- read_csv('presidents.csv') %>%
  cSplit("President Name,Years In Office,Number,Party", ",") %>% # split values into separate columns
  rename(president = "President Name,Years In Office,Number,Party_1", # rename columns
         years = "President Name,Years In Office,Number,Party_2",
         number = "President Name,Years In Office,Number,Party_3",
         party = "President Name,Years In Office,Number,Party_4") %>%
  separate(years, c("start_year", "end_year"), sep = "-") %>% # separate years values
  mutate(end_year = ifelse(president == "William H. Harrison", 1841, end_year)) %>% # fill in some misc missing values 
  mutate(end_year = ifelse(president == "Donald Trump", 2021, end_year)) %>%
  pivot_longer(cols = c(start_year, end_year), # pivot so that years are a column
               names_to = "start_end",
               values_to = "Year",
               values_transform = list("Year"=as.numeric)) %>%
  select(1, 3, 5) # select only rows we want

regress_set_b <- left_join(regress_set_a, presidents, by="Year") # add to dataset
regress_set_b

regress_set_b_1 <- regress_set_b %>% 
  fill(president, .direction = "up") %>%
  fill(party, .direction = "up") # ensure that all years have party and president values, not just start and end years
 
regress_set_b_2 <- regress_set_b_1 %>%
  fill(president, .direction = "down") %>% 
  fill(party, .direction = "down") # fill in where values are missing since no end year
regress_set_b_2
 
regress_set_b_2 <- regress_set_b_2[!duplicated(regress_set_b_2$Year, fromLast = T), ] # clear out any duplicated years by taking the last value of that year 
```

Add in population dataset; source: https://usafacts.org/data/topics/people-society/population-and-demographics/population-data/population/#chart-12818-0
 
```{r} 
population_source <- read_csv('population_usafacts.csv')

population <- population_source %>%
  filter(Years == "Population (People)") %>% # choose only population metric
  select(50:122) %>% # select only the years we are looking at
  pivot_longer(1:73, 
               names_to = "Year",
               values_to = "population",
               names_transform = list("Year"=as.numeric)) # pivot so that year is a column

regress_set_c <- left_join(regress_set_b_2, population, by="Year") # add to dataset
regress_set_c
``` 

Calculate population growth percentage
```{r}
regress_set_c$pop_growth <- vector("numeric", 71)

for (i in seq(nrow(regress_set_c))) {
  if (i == 1) {
    regress_set_c$pop_growth[i] <- 0
  } else {
    regress_set_c$pop_growth[i] <- ((regress_set_c$population[i] - regress_set_c$population[i - 1])/regress_set_c$population[i - 1])}
} 
```

# Exploratory Graphs

################# NOTES #################
consider: how to add in change years to identify possible trends??
most everything else is a year to year change except unemployment which is just a percentage...is this important?
add interpretations
Regression:
need to go back and review approaches to this
need to add logs???
#########################################

Alea: do you have tips on how to explore or present these things? I feel like there is potential for some fun visuals here, but want to make sure they are meaningful and relevant before going in too deep

Changes in minimum wage and average unemployment / population growth 
```{r}
regress_set_c %>%
  pivot_longer(cols = c(pop_growth, avg_unemploy),
                        names_to = "metric",
                        values_to = "values") %>%
  ggplot(aes(x = Year, y = values, group = metric, color = metric)) + 
    geom_line() +
    scale_color_manual(values = c("avg_unemploy" = "seagreen", "pop_growth" = "royalblue4")) +
    wage_theme
```

Changes in minimum wage and political party in power
```{r}
ggplot(regress_set_c, aes(x = Year, y = change, fill = party)) +
  geom_col() +
  scale_fill_manual(values = c("Democratic" = "seagreen", "Republican" = "royalblue4")) +
  wage_theme
         
# There seems to be no relationship between party and changes in minimum wage
```

Changes in minimum wage and unemployment
```{r}
regress_set_c %>%
  mutate(change = as.character(change)) %>%
  ggplot(aes(x = Year, y = avg_unemploy, fill = change)) + 
    geom_col() +
    scale_fill_manual(values = c("1" = "seagreen", "0" = "royalblue4")) +
    wage_theme
```

Fancy regression
```{r}
regress_set_c %>%
  mutate(change = as.character(change)) %>%
  ggplot() +
    aes(x = Year, y = avg_unemploy, group = change, color = change) +
    geom_point(color = "grey") +
    geom_smooth(method = "lm") +
    scale_color_manual(values = c("1" = "seagreen", "0" = "royalblue4")) +
    wage_theme
         
``` 

Changes in minimum wage and population
```{r}
regress_set_c %>%
  mutate(change = as.character(change)) %>%
  ggplot(aes(x = Year, y = pop_growth, fill = change)) + 
    geom_col() +
    scale_fill_manual(values = c("1" = "seagreen", "0" = "royalblue4")) +
    wage_theme
         
``` 

Regression attempt (come back to)
```{r}
min_wage_regress <- lm(data = regress_set_c, change ~ avg_unemploy + party + pop_growth)
summary(min_wage_regress)
```


