---
title: "Huber_Mokate_Project_4_27"
author: "Meghan Mokate and Jane Huber"
date: "4/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Minimum Wage is a piece of federal economic policy that sets a floor for wage work in the United
States. Originally brought into law during the Great Depression in the 1930’s, the aim was to create “minimum standard of living necessary for health, efficiency and general well-being, without substantially curtailing employment”. [source]

Ever since its creation, there has been vigorous debate across the US political spectrum as to what the value of a minimum wage should be or there should be one at all. As the “Fight for 15” political movement rises to prominence on the national level, some economists argue that $15/hour simply isn’t ambitious enough or a reflection of the economy as a whole. The Center for Economic and Policy Research (CEPR) claims, “Until 1968, the minimum wage not only kept pace with inflation, it rose in step with productivity growth. The logic is straightforward; we expect that
wages in general will rise in step with productivity growth. For workers at the bottom to share in the overall improvement in society’s living standards, the minimum wage should also rise with productivity.”
(https://cepr.net/this-is-what-minimum-wage-would-be-if-it-kept-pace-with-productivity)

This analysis will examine that claim through our own analysis of the numbers, looking to see what the value of the minimum wage would be if, from its inception, minimum wage was increased according to those two variables. Access to data to produce our own analysis is fairly straightforward, The US Bureau of Labor and Statistics (BLS) has tracked two key productivity measurements since 1948—a raw productivity rate (measuring output per hour) and a multifactor productivity rate (measuring output per unit of combined inputs) [(https://www.bls.gov/bls/productivity.htm)]. We will be using the raw productivity rate since we were able to access cleaned data

The value of inflation of the US dollar is important to factor in because it imbeds purchasing power of the dollar into what the minimum wage includes. For example, a study by Pew Research demonstrates that the minimum wage in 1973, $4.03/hr, has the same purchasing power of $23.68 in August 2018 — more than three times the actual minimum wage [https://www.pewresearch.org/facttank/2018/08/07/for-most-us-workers-real-wages-have-barely-budged-for-decades/].

Citations: 
https://cepr.net/this-is-what-minimum-wage-would-be-if-it-kept-pace-with-productivity/
https://www.bls.gov/bls/productivity.htm
https://www.pewresearch.org/fact-tank/2018/08/07/for-most-us-workers-real-wages-have-barely-
budged-for-decades 
https://www.forbes.com/advisor/personal-finance/minimumwage-debate/
https://democracyjournal.org/magazine/29/minimum-wage-catching-up-to-productivity/
https://www.nytimes.com/2020/06/18/business/coronavirus-minimum-wage-increase.html
https://www.cnn.com/interactive/2019/business/us-minimum-wage-by-year/index.html


# Guide to datasets

## Master Datasets
master_set = includes inflation adjusted
master_set_1 = includes productivity adjusted and inflation & productivity adjusted
master_set_2 = cleans unused columns priorly used for calculations
master_set_3 = pivoted file in order to graph for dollar disparities
master_set_4 = adds in percentage change columns
master_set_5 = pivoted file in order to graph for percentage disparities

## Regression Datasets
regress_set = master_set with dummy variable for years when minimum wage changed
regress_set_a = includes all and umemployment
regress_set_b = includes all and presidents for first step (do not use)
regress_set_b_1 = regress_set_b first step to cleanse (do not use)
regress_set_b_2 = regress_set_b cleansed
regress_set_c = includes all and population; this is the one will will use


# Import Libraries
```{r}
library(tidyverse) 
library(rvest)
library(janitor)
library(tinytex)
library(haven) # may use for regression / regression line (may not)
library(splitstackshape) # for cSplit
library(knitr)
library(extrafont)
library(blscrapeR) # to complete inflation-related calculations 
```

# PART ONE - CALCULATE MINIMUM WAGE

## Data Gathering, Cleaning, Tidying

Here, we gather and combine our datasets from several sources.
```{r data-gathering}

#Source: https://www.epi.org/productivity-pay-gap/
hourly_comp_net_prod_tb <- read_csv("Data_HourlyCompensation_NetProd.csv")

#Source: https://www.dol.gov/agencies/whd/minimum-wage/history/chart#fn4
min_wage_tb <- read_csv("Data_Year_Minwage.csv")

#Combine avg hourly compensation and net productivity datasets
total_minwage_comp_prod <- left_join(hourly_comp_net_prod_tb, min_wage_tb, by="Year")

#Fill in the federal minimum wage value; the original dataset only included information on when a law was passed to increase the federal minimum wage. Therefore, we need to fill in, year-by-year, what that value was.
total_minwage_comp_prod_r <- 
  total_minwage_comp_prod %>%
  fill(`Minimum Wage`, .direction = "down") 
total_minwage_comp_prod_r

# the 1948 and 1949 values of .40 were excluded, so add these in
total_minwage_comp_prod_r$`Minimum Wage`[is.na(total_minwage_comp_prod_r$`Minimum Wage`)] <- .40

#Retrieve data on annual inflation by scraping the USInflationCalculator website.
inflation_data_url <- "https://www.usinflationcalculator.com/inflation/consumer-price-index-and-annual-percent-changes-from-1913-to-2008/"
inflation_source <- inflation_data_url %>%
  html() %>%
  html_nodes(xpath='//*[@id="post-17"]/div/div[1]/table') %>%
  html_table()
inflation_source_1 <- inflation_source[[1]]

#Here, we create an annual inflation dataset by cleaning the data retrieved from the website.
inflation <- inflation_source_1 %>%
  row_to_names(row_number = 2) %>%
  select(1, 14) %>% # changed from column 16
  rename("Annual Inflation" = 3) %>%
  mutate(Year = as.double(Year), `Annual Inflation` = as.double(`Annual Inflation`))
inflation

#Finally, combine all datasets by year to create a master dataset that we will use for our full analysis
master_set <- left_join(total_minwage_comp_prod_r, inflation, by="Year")
master_set
```

## Calculate Values

Convert all Minimum Wage dollar values to 2018 dollars
```{r calculate values, warning=FALSE}
# start by putting our base year 
# $ amount CPI / year_0 CPI



```


Calculate expected minimum wage based off inflation
This means that the minimum wage in 1948 is adjusted solely for inflation annually and does not reflect congress-updated minimum wage values
```{r}
##### Alea: is this redundant of calculating minimum wage in 2018 dollars that we did above? Are we double counting the impact of inflation here?

# Change inflation value to a percentage
master_set$ann_inflat <- (master_set$`Annual Inflation`) / 100

# Create a vector to hold our inflation adjusted value
master_set$inflat_adj <- vector("numeric", 71)

for (i in seq(nrow(master_set))) { 
  if (i == 1) { # establish that the first value is just the 1948 wage in 2018 dollars
    master_set$inflat_adj[i] <- master_set$wage_18 # referencing new column we created to calculate minimum wage in 2018 values
  } else {
    master_set$inflat_adj[i] <- (master_set$inflat_adj[i - 1]*master_set$ann_inflat[i]) + master_set$inflat_adj[i-1] } # create calculation that references above adjusted inflation amount, multiplies it by the annual inflation number, and adds the inflation amount since this is how we can calculate these changes
}

# inflat_adj is now a new column in our dataset that shows 2018 minimum wage adjusted for inflation year to year

```


Calculate expected minimum wage based off productivity. This means that the minimum wage in 1948 is adjusted solely for productivity annually and does not reflect congress-updated minimum wage values.
```{r}
# adjust master_set to create a new column to calculate net productivity change so that we can calculate the year to year impact of net productivity on minimum wage
master_set_1 <- master_set %>% 
  mutate(net_prod_change = (`Net Productivity` - lag(`Net Productivity`, default = first(`Net Productivity`)))) %>% # current net productivity value minus previous
  mutate(net_prod_change = (net_prod_change / 100)) # adjust master_set net productivity change column to make it a percentage

# Create a vector to hold our productivity adjusted value
master_set_1$prod_adj <- vector("numeric", 71)

# Alea: should we set the base wage amount to the one adjusted for 2018 dollars, or the actual given ($0.40), this relates to the question of redundancy above
for (i in seq(nrow(master_set_1))) {
  if (i == 1) { # establish that the first value is just the 1948 wage in 2018 dollars
    master_set_1$prod_adj[i] <- master_set_1$wage_18 # referencing new column we created to calculate minimum wage in 2018 values -CHANGE IF REDUNDANT
  } else {
    master_set_1$prod_adj[i] <- (master_set_1$prod_adj[i - 1]*master_set_1$net_prod_change[i]) + master_set_1$prod_adj[i-1]} # create calculation that references above adjusted productivity amount, multiplies it by the annual productivity number, and adds the productivity amount since this is how we can calculate these changes
}
# prod_adj is now a new column in our dataset that shows 2018 minimum wage adjusted for productivity year to year

```

Calculate expected minimum wage based off inflation AND productivity. This means that the minimum wage in 1948 is adjusted for both inflation and productivity annually and does not reflect congress-updated minimum wage values.
```{r}
# Alea: We have two potential approaches here....

# The first is this: 
# Calculate inflation with productvity by multuplying inflation adjusted amount by 1 + net productivity change
master_set_1$inflat_prod_adj <- (master_set_1$inflat_adj * (1 + master_set_1$net_prod_change))
# CHANGE IF REDUNDANT

#The second is this (same as approach above):
# Create a vector to hold our productivity adjusted value
master_set_1$inflat_prod_adj <- vector("numeric", 71)

# Calculate net impact column that includes both inflation and net productivity change
master_set$net_impact <- master_set$ann_inflat + master_set$net_prod_change

for (i in seq(nrow(master_set_1))) {
  if (i == 1) {
    master_set_1$inflat_prod_adj[i] <- master_set_1$inflat_adj # # referencing new column we created to calculate inflation adjusted minimum wage - CHANGE IF REDUNDANT
  } else {
    master_set_1$inflat_prod_adj[i] <- (master_set_1$inflat_prod_adj[i - 1]*master_set_1$net_impact[i]) + master_set_1$inflat_prod_adj[i-1]} # create calculation that references above adjusted inflation & productivity amount, multiplies it by the annual net impact number (inflation & productivity percent change), and adds the inflation & productivity amount since this is how we can calculate these changes
}
```

Get rid of unused columns priorly used for calculations
```{r}
master_set_2 <- 
  master_set_1 %>% 
  select(-c(`Hourly Compensation`, `Annual Inflation`, `Net Productivity`))
```

##Pivot the dataset so that we can compare the following calculated minimum wage values: 
(1) Actual minimum wage (Congress)
(2) Inflation minimum wage
(3) Productivity minimum wage
(4) Inflation and productivity minimum wage
```{r}
master_set_3 <- 
  master_set_2 %>%
  pivot_longer(cols = c("Minimum Wage", inflat_adj, prod_adj, inflat_prod_adj),# CHANGE IF REDUNDANT
               names_to = "Type",
               values_to = "Dollars")

master_set_2

master_set_3
```

## Graph

Create a theme
```{r create-theme}
wage_theme <-
  theme_minimal() + 
  theme(
    text = element_text(family = "Georgia"), 
    plot.title = element_text(face = "bold", size = 16, color = "seagreen"), 
    plot.subtitle = element_text(face = "italic", size = 14, color = "royalblue4"), 
    axis.title = element_text(size = 14, face = "bold", color = "seagreen"), 
    axis.text = element_text(size = 12, color = "navyblue"), 
    axis.line = element_blank(), 
    axis.ticks = element_blank(), 
    axis.text.x = element_text(angle = 90), 
    legend.title = element_text(size = 12, face = "bold", color = "seagreen"), 
    legend.text = element_text(size = 12, color = "navyblue")) 
```

## Graph Values for Dollar Disparities in Actual and Calculated Minimum Wage Changes Year over Year
```{r graph-values, fig.width=9}
wage_values_graph <- 
  master_set_3 %>% 
  ggplot() +
  geom_path(aes(x = Year, y=Dollars, color= Type)) +
  labs(title = "$ Disparity in Actual and Calculated Minimum Wage Changes") +
  wage_theme
wage_values_graph

# thicken lines
```


Find actual increase percent value for minimum wage.
```{r other-calculations, warnings = FALSE}
# update dataset name for new calculations
master_set_4 <- master_set_2

# Create a vector to hold our actual increase percent value
master_set_4$increase_act <- vector("numeric", 71)

for (i in seq(nrow(master_set_4))) {
  if (i == 1) {
    master_set_4$increase_act[i] <- 0
  } else {
    master_set_4$increase_act[i] <- ((master_set_4$`Minimum Wage`[i] - master_set_4$`Minimum Wage`[i - 1])/master_set_4$`Minimum Wage`[i - 1])} # CHANGE IF REDUNDANT (NOT) to wage_18
}
# create calculation to find percentage change from one year to the next year
```

Find accumulated calculated increase percent value for minimum wage.
```{r}
# Create a vector to hold our accumulated CALCULATED increase percent value 
master_set_4$accum_incr_calc <- vector("numeric", 71)

for (i in seq(nrow(master_set_4))) {
  if (i == 1) {
    master_set_4$accum_incr_calc[i] <- 0
  } else {
    master_set_4$accum_incr_calc[i] <- ((master_set_4$net_impact[i] + master_set_4$accum_incr_calc[i - 1]))} # add net income to running total year by year
}
```

Find accumulated actual increase percent value for minimum wage.
```{r}
# Create a vector to hold our accumulated ACTUAL increase percent value 
master_set_4$accum_incr_act <- vector("numeric", 71)

for (i in seq(nrow(master_set_4))) {
  if (i == 1) {
    master_set_4$accum_incr_act[i] <- 0
  } else {
    master_set_4$accum_incr_act[i] <- ((master_set_4$increase_act[i] + master_set_4$accum_incr_act[i - 1]))}
}
```

## Pivot the dataset so that we can compare the following calculated minimum wage percentage values:
(1) Inflation Percentage Year to Year
(2) Productivity Change Percentage Year to Year
(3) Calculated Net Impact Percentage Year to Year
(4) Accumulated Calculated Percentage Change Running Total of Years
(5) Accumulated Actual Percentage Change Running Total of Years

```{r}
master_set_5 <- 
  master_set_4 %>%
  pivot_longer(cols = c(ann_inflat, net_prod_change, net_impact, accum_incr_calc, accum_incr_act),
               names_to = "Increase Type",
               values_to = "Increase Amount")
master_set_5
```

## Graph Values for Percentage Disparities in Actual and Calculated Minimum Wage Changes Year over Year
```{r, fig.width=9}
ggplot() +  
  geom_path(data = master_set_5, aes(x = Year, y = `Increase Amount`, color = `Increase Type`)) +
  labs(title = "% Disparity in Actual and Calculated Minimum Wage Changes") +
  wage_theme
```

# PART TWO - IMPACTFUL FACTORS

## Regression Datasets and Analysis

Create dummy variable for when there was a change in minimum wage
```{r}
regress_set <- master_set %>%
  mutate(change = ifelse(increase_act == 0, 0, 1))
```

## Add in regression datasets: 
(1) unemployment rate
(2) president / party
(3) population / population growth

Unemployment dataset; source: https://data.bls.gov/pdq/SurveyOutputServlet
```{r}
unemploy <- read_csv('unemploy.csv') %>%
  row_to_names(row_number = 11) %>% # make values in row 11 the column titles
  mutate_if(is.character,as.numeric) %>% #change all values to numeric
  rowwise() %>% # prepare to take mean of values
  mutate(avg_unemploy = mean(c_across(2:13))) %>% # create column of mean to get avg_unemploy
  select(1, 14) %>% # select only year and average column
  mutate(avg_unemploy = (avg_unemploy / 100)) # make this into a percentage

regress_set_a <- left_join(regress_set, unemploy, by="Year") # add to dataset
regress_set_a
```

Presidential party for each year dataset; source: https://github.com/awhstin/Dataset-List/blob/master/presidents.csv
```{r}
presidents <- read_csv('presidents.csv') %>%
  cSplit("President Name,Years In Office,Number,Party", ",") %>% # split values into separate columns
  rename(president = "President Name,Years In Office,Number,Party_1", # rename columns
         years = "President Name,Years In Office,Number,Party_2",
         number = "President Name,Years In Office,Number,Party_3",
         party = "President Name,Years In Office,Number,Party_4") %>%
  separate(years, c("start_year", "end_year"), sep = "-") %>% # separate years values
  mutate(end_year = ifelse(president == "William H. Harrison", 1841, end_year)) %>% # fill in some misc missing values 
  mutate(end_year = ifelse(president == "Donald Trump", 2021, end_year)) %>%
  pivot_longer(cols = c(start_year, end_year), # pivot so that years are a column
               names_to = "start_end",
               values_to = "Year",
               values_transform = list("Year"=as.numeric)) %>%
  select(1, 3, 5) # select only rows we want

regress_set_b <- left_join(regress_set_a, presidents, by="Year") # add to dataset
regress_set_b

regress_set_b_1 <- regress_set_b %>% 
  fill(president, .direction = "up") %>%
  fill(party, .direction = "up") # ensure that all years have party and president values, not just start and end years
 
regress_set_b_2 <- regress_set_b_1 %>%
  fill(president, .direction = "down") %>% 
  fill(party, .direction = "down") # fill in where values are missing since no end year
regress_set_b_2
 
regress_set_b_2 <- regress_set_b_2[!duplicated(regress_set_b_2$Year, fromLast = T), ] # clear out any duplicated years by taking the last value of that year 
```

Add in population dataset; source: https://usafacts.org/data/topics/people-society/population-and-demographics/population-data/population/#chart-12818-0
 
```{r} 
population_source <- read_csv('population_usafacts.csv')

population <- population_source %>%
  filter(Years == "Population (People)") %>% # choose only population metric
  select(50:122) %>% # select only the years we are looking at
  pivot_longer(1:73, 
               names_to = "Year",
               values_to = "population",
               names_transform = list("Year"=as.numeric)) # pivot so that year is a column

regress_set_c <- left_join(regress_set_b_2, population, by="Year") # add to dataset
regress_set_c
``` 

Calculate population growth percentage
```{r}
regress_set_c$pop_growth <- vector("numeric", 71)

for (i in seq(nrow(regress_set_c))) {
  if (i == 1) {
    regress_set_c$pop_growth[i] <- 0
  } else {
    regress_set_c$pop_growth[i] <- ((regress_set_c$population[i] - regress_set_c$population[i - 1])/regress_set_c$population[i - 1])}
} 
```

# Exploratory Graphs

################# NOTES #################
consider: how to add in change years to identify possible trends??
most everything else is a year to year change except unemployment which is just a percentage...is this important?
add interpretations
Regression:
need to go back and review approaches to this
need to add logs???
#########################################

Alea: do you have tips on how to explore or present these things? I feel like there is potential for some fun visuals here, but want to make sure they are meaningful and relevant before going in too deep

Changes in minimum wage and average unemployment / population growth 
```{r}
regress_set_c %>%
  pivot_longer(cols = c(pop_growth, avg_unemploy),
                        names_to = "metric",
                        values_to = "values") %>%
  ggplot(aes(x = Year, y = values, group = metric, color = metric)) + 
    geom_line() +
    scale_color_manual(values = c("avg_unemploy" = "seagreen", "pop_growth" = "royalblue4")) +
    wage_theme
```

Changes in minimum wage and political party in power
```{r}
ggplot(regress_set_c, aes(x = Year, y = change, fill = party)) +
  geom_col() +
  scale_fill_manual(values = c("Democratic" = "seagreen", "Republican" = "royalblue4")) +
  wage_theme
         
# There seems to be no relationship between party and changes in minimum wage
```

Changes in minimum wage and unemployment
```{r}
regress_set_c %>%
  mutate(change = as.character(change)) %>%
  ggplot(aes(x = Year, y = avg_unemploy, fill = change)) + 
    geom_col() +
    scale_fill_manual(values = c("1" = "seagreen", "0" = "royalblue4")) +
    wage_theme
```

Fancy regression
```{r}
regress_set_c %>%
  mutate(change = as.character(change)) %>%
  ggplot() +
    aes(x = Year, y = avg_unemploy, group = change, color = change) +
    geom_point(color = "grey") +
    geom_smooth(method = "lm") +
    scale_color_manual(values = c("1" = "seagreen", "0" = "royalblue4")) +
    wage_theme
         
``` 

Changes in minimum wage and population
```{r}
regress_set_c %>%
  mutate(change = as.character(change)) %>%
  ggplot(aes(x = Year, y = pop_growth, fill = change)) + 
    geom_col() +
    scale_fill_manual(values = c("1" = "seagreen", "0" = "royalblue4")) +
    wage_theme
         
``` 

Regression attempt (come back to)
```{r}
min_wage_regress <- lm(data = regress_set_c, change ~ avg_unemploy + party + pop_growth)
summary(min_wage_regress)
```
